Computational methods in landscape ecology

### 2024-08-30 Manuscript.qmd
Process to connectivity section

---
title: "Computational methods in landscape ecology"
keywords: keyword1, keyword2
author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
    attributes:
      - corresponding: boolean
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affiliations:
      - name: Adam Mickiewicz University, Poznan, Poland
      - name: University of Münster, Münster, Germany
  - name: Marti Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affiliations:
      - name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
  - name: Rafael Schouten
    orcid: 0000-0002-8380-0884
    email: rafael.schouten@sund.ku.dk
    affiliations:
      - name: Globe Institute, University of Copenhagen, Copenhagen, Denmark
format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
papersize: a4paper
fontsize: 10pt
geometry: margin=25mm
indent: 2m
linestretch: 1.5
bibliography: CLER.bib
csl: springer-basic-author-date.csl
---

>>>>

---
title: "Computational methods in landscape ecology"
keywords: keyword1, keyword2
author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
    attributes:
      - corresponding: boolean
  - name: Jakub Nowosad
    orcid: 0000-0002-1057-3721
    email: nowosad.jakub@gmail.com
    affiliations:
      - name: Adam Mickiewicz University, Poznan, Poland
      - name: University of Münster, Münster, Germany
  - name: Marti Bosch
    orcid: 0000-0001-8735-9144
    email: marti.bosch@epfl.ch
    affiliations:
      - name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
  - name: Martin Jung
    orcid: 0000-0002-7569-1390
    email: jung@iiasa.ac.at
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
papersize: a4paper
fontsize: 10pt
geometry: margin=25mm
indent: 2m
linestretch: 1.5
bibliography: CLER.bib
csl: springer-basic-author-date.csl
---

### 2024-08-30 Manuscript.tex
Process to connectivity section

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Computational methods in landscape ecology},
  pdfauthor={Maximilian H.K. Hesselbarth; Jakub Nowosad; Marti Bosch; Martin Jung; Rafael Schouten},
  pdfkeywords={keyword1, keyword2},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

  \author{Maximilian H.K. Hesselbarth}
            \affil{%
                  International Institute for Applied Systems Analysis,
                  Biodiversity, Ecology, and Conservation Group,
                  Laxenburg, Austria
              }
        \author{Jakub Nowosad}
            \affil{%
                  Adam Mickiewicz University, Poznan, Poland
              }
          \affil{%
                  University of Münster, Münster, Germany
              }
        \author{Marti Bosch}
            \affil{%
                  École polytechnique fédérale de Lausanne, Lausanne,
                  Switzerland
              }
        \author{Martin Jung}
            \affil{%
                  International Institute for Applied Systems Analysis,
                  Biodiversity, Ecology, and Conservation Group,
                  Laxenburg, Austria
              }
        \author{Rafael Schouten}
            \affil{%
                  Globe Institute, University of Copenhagen, Copenhagen,
                  Denmark
              }

@@

Connectivity describes how movement among resources patches is
facilitated or hindered by landscape features such as types, amounts,
and arrangement of LULC (Taylor et al. 1993, 2006). While structural
connectivity focuses on physical and spatial relationships among
resource patches, functional connectivity also includes the movement or
flow of individuals and populations through the landscape (Taylor et al.
2006). Connectivity underpins ecosystem functionality and maintains
biodiversity, and plays a large role in many conservation actions
(Correa Ayram et al. 2016; Hilty et al. 2020). Thus, in the past
multiple approaches how to measure connectivity were developed (Simpkins
et al. 2018; Keeley et al. 2021).

@@

\begin{itemize}
\tightlist
\item
  Many traditional lsm measure connectivity, e.g, ENN
\item
  Recently, also focus in within-patch connectivity
  (https://doi.org/10.1007/s10980-019-00881-0)
\end{itemize}

@@

https://doi.org/10.1111/2041-210X.12750
https://doi.org/10.1007/s10980-019-00789-9
https://doi.org/10.1007/s10980-022-01504-x
https://doi.org/10.1007/s10980-022-01469-x

\subsection{Circuit theory-based}\label{circuit-theory-based}

https://circuitscape.org https://doi.org/10.1111/2041-210X.12689
https://doi.org/10.1016/j.ecolind.2016.03.032

\subsection{Least-cost paths (LCPs)}\label{least-cost-paths-lcps}

https://doi.org/10.1016/S0169-2046(02)00242-6
https://doi.org/10.1007/s40823-016-0006-9

@@

https://doi.org/10.1007/s10980-020-00998-7
https://doi.org/10.1007/s10980-020-01164-9

@@

https://doi.org/10.1890/14-1690.1
https://doi.org/10.1016/j.ecolind.2016.03.032
https://doi.org/10.1007/s10980-019-00935-3
https://doi.org/10.1126/science.abg1780

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{1}
\bibitem[\citeproctext]{ref-Angel2010}
Angel S, Parent J, Civco DL (2010) Ten compactness properties of
circles: Measuring shape in geography. Canadian Geographies /
G{é}ographies canadiennes 54:441--461.
\url{https://doi.org/10.1111/j.1541-0064.2009.00304.x}

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Computational methods in landscape ecology},
  pdfauthor={Maximilian H.K. Hesselbarth; Jakub Nowosad; Marti Bosch; Martin Jung},
  pdfkeywords={keyword1, keyword2},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

  \author{Maximilian H.K. Hesselbarth}
            \affil{%
                  International Institute for Applied Systems Analysis,
                  Biodiversity, Ecology, and Conservation Group,
                  Laxenburg, Austria
              }
        \author{Jakub Nowosad}
            \affil{%
                  Adam Mickiewicz University, Poznan, Poland
              }
          \affil{%
                  University of Münster, Münster, Germany
              }
        \author{Marti Bosch}
            \affil{%
                  École polytechnique fédérale de Lausanne, Lausanne,
                  Switzerland
              }
        \author{Martin Jung}
            \affil{%
                  International Institute for Applied Systems Analysis,
                  Biodiversity, Ecology, and Conservation Group,
                  Laxenburg, Austria
              }

@@

Connectivity describes how movement among resources patches is
facilitated or hindered by landscape features such as types, amounts,
and arrangement of LULC (Taylor et al. 1993, 2006). While structural
connectivity focuses on physical and spatial relationships among
resource patches, functional connectivity also includes the movement or
flow of individuals and populations through the landscape (Taylor et al.
2006). Connectivity underpins ecosystem functionality and maintains
biodiversity and populations, and plays a large role in many
conservation actions (Correa Ayram et al. 2016; Hilty et al. 2020).
Thus, in the past multiple approaches how to measure connectivity were
developed (Simpkins et al. 2018; Keeley et al. 2021).

@@

There are several landscape metrics that are traditionally used to
measure landscape connectivity. For example, the habitat amount within a
buffer zone of focal patches, the nearest-neighbor distance, or the
connectance index (Moilanen and Nieminen 2002; Prugh 2009; Spanowicz and
Jaeger 2019). However, these metrics are frequently criticised for being
to simple and especially the link to functional connectivity is often
ambiguous. Thus, more functionality needs to be included using
additional ecological meaningful metrics, such as the core area (e.g.,
Kupfer 2012). More recently there is also an increasing focus in
within-patch connectivity using metrics such as the effective mesh size
in addition to the usual measure between-patch connectivity (Spanowicz
and Jaeger 2019).

@@

Resistance surfaces are one of the most popular approaches to model
connectivity and developed from simple isolation-by-distance approaches.
The approach uses environmental layers and biodiversity data or expert
opinions to estimate the cost and of movement through landscapes (Zeller
et al. 2012). The latest developments of resistance surfaces have
recently been reviewed (Dutta et al. 2022). The resistance surface can
be used to model movement based on different approaches such as
leas-cost paths (Adriaensen et al. 2003; Etherington 2016) or circuit
theory (McRae et al. 2008). However, resistance surfaces are not free of
critique including missing spatial-temporal variability or context
dependency (Unnithan Kumar et al. 2022b; Unnithan Kumar and Cushman
2022). Thus, recent developments include the combination with more
processes-based modelling approaches (Schumaker and Brookes 2018; Bocedi
et al. 2021; Unnithan Kumar et al. 2022a), and modelling of dynamic
landscape connectivity (Zeller et al. 2020).

@@

@@

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{1}
\bibitem[\citeproctext]{ref-Adriaensen2003}
Adriaensen F, Chardon JP, De Blust G, et al (2003) The application of
{``least-cost''} modelling as a functional landscape model. Landscape
and Urban Planning 64:233--247.
\url{https://doi.org/10.1016/S0169-2046(02)00242-6}

\bibitem[\citeproctext]{ref-Angel2010}
Angel S, Parent J, Civco DL (2010) Ten compactness properties of
circles: Measuring shape in geography. Canadian Geographies /
G{é}ographies canadiennes 54:441--461.
\url{https://doi.org/10.1111/j.1541-0064.2009.00304.x}

@@

\bibitem[\citeproctext]{ref-Bocedi2021}
Bocedi G, Palmer SCF, Malchow A-K, et al (2021) RangeShifter 2.0: An
extended and enhanced platform for modelling spatial eco-evolutionary
dynamics and species' responses to environmental changes. Ecography
44:1453--1462. \url{https://doi.org/10.1111/ecog.05687}

@@

\bibitem[\citeproctext]{ref-Dutta2022}
Dutta T, Sharma S, Meyer NFV, et al (2022) An overview of computational
tools for preparing, constructing and using resistance surfaces in
connectivity research. Landscape Ecology 37:2195--2224.
\url{https://doi.org/10.1007/s10980-022-01469-x}

\bibitem[\citeproctext]{ref-Etherington2016}
Etherington TR (2016) Least-cost modelling and landscape ecology:
Concepts, applications, and opportunities. Current Landscape Ecology
Reports 1:40--53. \url{https://doi.org/10.1007/s40823-016-0006-9}

@@

\bibitem[\citeproctext]{ref-McRae2008}
McRae B, Dickson BG, Keitt TH, Shah VB (2008) Using circuit theory to
model connectivity in ecology, evolution, and conservation. Ecology
89:2712--2724. \url{https://doi.org/10.1890/07-1861.1}

@@

\bibitem[\citeproctext]{ref-Moilanen2002}
Moilanen A, Nieminen M (2002) Simple connectivity measures in spatial
ecology. Ecology 83:1131--1145.
\url{https://doi.org/10.1890/0012-9658(2002)083\%5B1131:SCMISE\%5D2.0.CO;2}

@@

\bibitem[\citeproctext]{ref-Prugh2009}
Prugh LR (2009) An evaluation of patch connectivity measures. Ecological
Applications 19:1300--1310. \url{https://doi.org/10.1890/08-1524.1}

@@

\bibitem[\citeproctext]{ref-Spanowicz2019}
Spanowicz AG, Jaeger JAG (2019) Measuring landscape connectivity: On the
importance of within-patch connectivity. Landscape Ecology
34:2261--2278. \url{https://doi.org/10.1007/s10980-019-00881-0}

@@

\bibitem[\citeproctext]{ref-UnnithanKumar2022a}
Unnithan Kumar S, Cushman SA (2022) Connectivity modelling in
conservation science: A comparative evaluation. Scientific Reports
12:16680. \url{https://doi.org/10.1038/s41598-022-20370-w}

\bibitem[\citeproctext]{ref-UnnithanKumar2022b}
Unnithan Kumar S, Kaszta Z, Cushman SA (2022a) Pathwalker: A new
individual-based movement model for conservation science and
connectivity modelling. ISPRS International Journal of Geo-Information
11:329. \url{https://doi.org/10.3390/ijgi11060329}

\bibitem[\citeproctext]{ref-UnnithanKumar2022}
Unnithan Kumar S, Turnbull J, Hartman Davies O, et al (2022b) Moving
beyond landscape resistance: Considerations for the future of
connectivity modelling and conservation science. Landscape Ecology
37:2465--2480. \url{https://doi.org/10.1007/s10980-022-01504-x}

@@

\bibitem[\citeproctext]{ref-Zeller2012}
Zeller KA, McGarigal K, Whiteley AR (2012) Estimating landscape
resistance to movement: A review. Landscape Ecology 27:777--797.
\url{https://doi.org/10.1007/s10980-012-9737-0}

\bibitem[\citeproctext]{ref-Zeller2020}
Zeller K, Lewison R, Fletcher R, et al (2020) Understanding the
importance of dynamic landscape connectivity. Land 9:303.
\url{https://doi.org/10.3390/land9090303}

### 2024-08-30 Sections/03-connectivity.qmd
Process to connectivity section

Connectivity describes how movement among resources patches is facilitated or hindered by landscape features such as types, amounts, and arrangement of LULC [@Taylor1993; @Taylor2006].
While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006].
Connectivity underpins ecosystem functionality and maintains biodiversity, and plays a large role in many conservation actions [@CorreaAyram2016; @Hilty2020].
Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018; @Keeley2021].

@@

- Many traditional lsm measure connectivity, e.g, ENN
- Recently, also focus in within-patch connectivity (https://doi.org/10.1007/s10980-019-00881-0)

@@

https://doi.org/10.1111/2041-210X.12750
https://doi.org/10.1007/s10980-019-00789-9
https://doi.org/10.1007/s10980-022-01504-x
https://doi.org/10.1007/s10980-022-01469-x

## Circuit theory-based

https://circuitscape.org
https://doi.org/10.1111/2041-210X.12689
https://doi.org/10.1016/j.ecolind.2016.03.032

##  Least-cost paths (LCPs)

https://doi.org/10.1016/S0169-2046(02)00242-6
https://doi.org/10.1007/s40823-016-0006-9

@@

https://doi.org/10.1007/s10980-020-00998-7
https://doi.org/10.1007/s10980-020-01164-9

@@

https://doi.org/10.1890/14-1690.1
https://doi.org/10.1016/j.ecolind.2016.03.032
https://doi.org/10.1007/s10980-019-00935-3
https://doi.org/10.1126/science.abg1780

>>>>

Connectivity describes how movement among resources patches is facilitated or hindered by landscape features such as types, amounts, and arrangement of LULC [@Taylor1993; @Taylor2006].
While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006].
Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016; @Hilty2020].
Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018; @Keeley2021].

<!-- 
Connectivity modelling in conservation science: https://doi.org/10.1038/s41598-022-20370-w
-->

@@

There are several landscape metrics that are traditionally used to measure landscape connectivity.
For example, the habitat amount within a buffer zone of focal patches, the nearest-neighbor distance, or the connectance index [@Moilanen2002; @Prugh2009; @Spanowicz2019].
However, these metrics are frequently criticised for being to simple and especially the link to functional connectivity is often ambiguous. 
Thus, more functionality needs to be included using additional ecological meaningful metrics, such as the core area [e.g., @Kupfer2012].
More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019].

@@

Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012].
The latest developments of resistance surfaces have recently been reviewed [@Dutta2022].
The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003; @Etherington2016] or circuit theory [@McRae2008].
However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022; @UnnithanKumar2022a].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018; @Bocedi2021; @UnnithanKumar2022b], and modelling of dynamic landscape connectivity [@Zeller2020]. <!-- MH: Not complete yet -->

<!--
lscorridors: https://doi.org/10.1111/2041-210X.12750
Complementary landscape genetics and habitat modelling approaches: https://doi.org/10.1007/s10980-019-00789-9
https://doi.org/10.1111/2041-210X.12689
https://doi.org/10.1016/j.ecolind.2016.03.032
-->

@@

<!--
https://doi.org/10.1007/s10980-020-00998-7
https://doi.org/10.1007/s10980-020-01164-9
-->

@@

<!--
https://doi.org/10.1890/14-1690.1
https://doi.org/10.1016/j.ecolind.2016.03.032
https://doi.org/10.1007/s10980-019-00935-3
https://doi.org/10.1126/science.abg1780
-->

### 2024-08-30 Sections/05-simulations.qmd
Process to connectivity section

<!-- Person responsible: Hesselbarth, Schouten -->

@@

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or binary space partitioning [@Etherington2022a].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [NLMR\; @Sciaini2018], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->

>>>>

<!-- Person responsible: Hesselbarth -->

@@

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or binary space partitioning [@Etherington2022a].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [ c], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->
